Predicting _C. difficile_ Infection Severity from the Taxonomic Composition of the Gut Microbiome

### 2023-06-05 paper/paper.qmd
shaping up discussion

There is not currently a consensus definition of CDI severity.
Some scoring systems such as the IDSA severity score focus on laboratory data
available during the course of CDI, while others focus on adverse outcomes of
CDI at 30 days after diagnosis [@ressler_defining_2021].
We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
The IDSA definition of severe CDI is based on laboratory values collected on
the day of diagnosis, with a case being severe if serum creatinine level is
greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
The remaining definitions focus on the occurrence of adverse outcomes, which
may be more clinically relevant.
The all-cause severity definition defines a case as severe if ICU admission,
colectomy, or death occurs within 30 days of CDI diagnosis, regardless of the
cause of the adverse event.
The attributable severity definition is based on disease-related complications
defined by the CDC, where an adverse event of ICU admission, colectomy, or death
occurs within 30 days of CDI diagnosis, and the adverse event is determined to
be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
Finally, we defined a pragmatic severity definition that makes use of the
attributable definition when available and falls back to the all-cause
definition when chart review has not been completed, allowing us to use as many
samples as we have available while taking physicians' expert opinions into
account where possible (@fig-flowchart B).
We trained ML models to classify (in the case of the IDSA definition) or predict
(in the case of the three other definitions) severity for each definition and
determined how well OTU-based models perform for each of them.

@@

Performance.
full datasets for best models possible for clinical application.
intersection dataset for comparing severity definitions.

The IDSA definition is known to be a poor predictor of adverse outcomes [@stevens_validation_2020],
however, it is easy to collect.

Discuss important OTUs. which ones concord with literature, which ones may be new.
Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and
severity has been well-documented in prior studies, thus its importance and
increase in abundance for severe cases is not surprising
[@schubert_microbiome_2014;
@antharam_intestinal_2013;
@berkell_microbiota-based_2021;
@lesniak_gut_2022].
For many of the top OTUs, there is wide variance in importance, perhaps due to
the imbalanced nature of the severity outcomes.
e.g. pseudomonas super high variance for attributable, highest point more 
important than any other OTU, but minimum CI crosses zero -- some data splits
are unlucky? could be more important if more data available?
Abundance data are sparse, likely due to these patients being on antibiotics.
Really showcases importance of having as many samples as possible when data are
sparse and the outcome is low prevalence.
we do not know which antibiotics were prescribed to treat these CDI cases, nor
which antibiotics patients may have taken prior to the CDI.
differences in microbiota between patients may be due to different abx...
different antibiotics have been shown to create different forms of dysbiotic
microbiota [@berkell_microbiota-based_2021] and
differential cdi clearance [@tomkovich_initial_2020].
vanc-resistant enterococcus new nosocomial alliance [@poduval_clostridium_2000].

Compare to EHR-based models.
tbh gut microbiome may only play a small role.
in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be -- can be embedded for automated scores.
further work needed to characterize the costs and benefits.

Models to guide treatment options. In the case of low-risk and non-invasive treatments
such as choosing between oral antibiotics, a higher number of false positives may
be tolerable as long as treatment cost is not unbearably high. However, for highly
invasive and irreversibly treatments such as colectomy, false positives must be minimized.
Cite studies saying fidaxomicin is cost-effective relative to vancomycin -
mentioned by @johnson_clinical_2021, e.g. @jiang_budget_2022.

It's not enough for models to perform well to justify deploying them in a
clinical setting; benefit over current practices must be shown.
do no harm [@wiens_no_2019].
Estimating the NNB contextualizes model performance within clinical reality.
Amplicon sequencing is not typically performed for CDI patients, but if there is
clinical value to be gained by implementing OTU-based models, routinely
sequencing and profiling the microbial communities of CDI patients could be
justified.
resistance to vancomycin is increasing in staph, cdi, and enterococci -- even
more important to find alternate treatments.
analysis of clinical value could be extended to include costs as well as include
other treatments if evidence of severity prevention emerges.
bezlotoxumab shown to prevent systemic organ damage in a mouse study [@mileto_bezlotoxumab_2022].

Models predicting the pragmatic definition yielded the best NNS.
While the attributable definition had a worse NNS for our OTU-based models, it
did not perform worse than the prior curated model, and it is the most
clinically relevant as physician chart review increases confidence that
positively-labelled severe outcomes are due to the CDI rather than other causes.

@@

>>>>

There is not currently a consensus definition of CDI severity.
Some scoring systems leverage clinical data available during the course of CDI,
while others focus on adverse outcomes of CDI at 30 days after diagnosis
[@ressler_defining_2021].
We explored four different ways to define CDI cases as severe or not (@fig-flowchart).
The "IDSA" definition of severe CDI is based on laboratory values collected on
the day of diagnosis, with a case being severe if serum creatinine level is
greater than or equal to $1.5 mg/dL$ and the white blood cell count is greater
than or equal to $15 k/\mu L$ [@mcdonald_clinical_2018].
However, the IDSA score is known to be a poor predictor of adverse outcomes
[@stevens_validation_2020], although it is straightforward to collect.
The remaining definitions we employed focus on the occurrence of adverse
outcomes, which may be more clinically relevant.
The "attributable" severity definition is based on disease-related complications
defined by the CDC, where an adverse event of ICU admission, colectomy, or death
occurs within 30 days of CDI diagnosis, and the adverse event is determined to
be attributable to the CDI by physician chart review [@mcdonald_recommendations_2007].
However, physician chart review is time-consuming and has not been completed for
all cases, so we defined "all-cause" severity where a case is severe if an
adverse event occurs within 30 days of the diagnosis regardless of the cause of
the adverse event.
Finally, we defined a "pragmatic" severity definition that makes use of the
attributable definition when available and falls back to the all-cause
definition when chart review has not been completed, allowing us to use as many
samples as we have available while taking physicians' expert opinions into
account where possible (@fig-flowchart B).
We trained ML models to classify (in the case of the IDSA definition) or predict
(in the case of the three other definitions) severity for each definition and
determined how well OTU-based models perform for each of them.

@@

We trained ML models based on gut microbial communities on the day of CDI
diagnosis to predict CDI severity according to four different severity definitions.
The purpose of using the full datasets is to train the best models possible
given the constraints, while using the intersection dataset allows for comparing
severity definitions.
We found that models predicting pragmatic severity with as much data as
available performed best, and that performance dropped substantially when 
reducing to the intersection datasets for all definitions likely due to the
particularly imbalanced nature of the all-cause and attributable definitions.
These results demonstrate the importance of using as many samples as possible when
data are sparse and the outcome is low prevalence, as well as the need to
incorporate physician's expertise when possible.

Permutation feature importance revealed patterns of important bacteria that
concord with the literature.
Enrichment of _Enterococcus_ and _Lactobacillus_ in _C. difficile_ infection and
severity has been well-documented in prior studies, thus its importance and
increase in abundance for severe cases is not surprising
[@schubert_microbiome_2014;
@antharam_intestinal_2013;
@berkell_microbiota-based_2021;
@lesniak_gut_2022].
For many of the top OTUs, there is a wide range in importance.
Notably, the OTU represented by _Pseudomonas_ had wide variance in importance
for the full dataset in models predicting attributable severity, with the
maximum point more important than any other OTU yet a minimum below zero.
However, for the intersection dataset, this OTU was removed due to having
near-zero variance.
The presence of _Pseudomonas_ was thus informative in a small number of 
patient samples, but not in others, and these samples were lost in the
intersection dataset.
Overall the abundance data are sparse, as these patients were likely on
antibiotics prior to CDI onset and may have begun antibiotic treatment prior to
the initial stool sample collection.

The full pragmatic severity model performed just as well as a prior
EHR-based model trained on the day of diagnosis, demonstrating the potential
utility of OTU-based models.
However, it is not enough for models to perform well to justify deploying them
in a clinical setting; benefit over current practices must be shown.
We calculated the number needed to screen for our OTU-based models and compared
these to EHR-based and curated models trained two days after the diagnosis, and
found the OTU-based pragmatic severity model outperformed the curated models but
not the EHR models.
While the attributable definition had a worse NNS for the OTU-based models, it
did not perform worse than the prior curated model, and it may be the most
clinically relevant as physician chart review increases confidence that
positively-labelled severe outcomes are due to the CDI rather than other causes.

Although no known treatment options are proven to reduce the risk of severe CDI
outcomes, fidaxomicin is promising due to its improved time to resolution and
reduced recurrence.
Fidaxomicin is also attractive as a preferential antibiotic option as
vancomycin-resistant enterococcus is on the rise and enterococci are known to
worsen CDI [@poduval_clostridium_2000;@smith_enterococci_2022].
We extended our analysis of clinical value to incorporate the number needed to
treat for fidaxomicin alongside the predictive models in order to calculate the
number needed to benefit.
The NNB contextualizes model performance within clinical reality.

A more robust analysis of clinical value would further consider the cost of
treatment options versus the savings of averting severe outcomes, as economic
disparities are a major barrier to treatment in the US.
Cost-benefit analyses have found that fidaxomicin is more cost-effective than
vancomycin largely due to the reduced risk of recurrence [@johnson_clinical_2021;@jiang_budget_2022].

As patient age is already the most significant risk factor, the gut microbiome
may only a minor part in influencing CDI severity.
However, deploying models incorporating both microbial factors and patient
factors from the EHR could be worth the time and costs if
performance improvements were shown translate to clinically significant improvements in
patient outcomes.
in practice, EHR-based models are easier and less costly to deploy than an OTU-based model would be -- can be embedded for automated scores.
further work needed to characterize the costs and benefits.

Amplicon sequencing is not typically performed for CDI patients, but if there is
clinical value to be gained by implementing OTU-based models, routinely
sequencing and profiling the microbial communities of CDI patients could be
justified.

Models predicting the pragmatic definition yielded the best NNS.
Future work is needed to determine whether incorporating both microbial features
and EHRs would result in a clinically meaningful improvement in performance over
deploying EHR models alone.

@@

<!--
we do not know which antibiotics were prescribed to treat these CDI cases, nor
which antibiotics patients may have taken prior to the CDI.
differences in microbiota between patients may be due to different abx...
different antibiotics have been shown to create different forms of dysbiotic
microbiota [@berkell_microbiota-based_2021] and
differential cdi clearance [@tomkovich_initial_2020].

Models to guide treatment options. In the case of low-risk and non-invasive treatments
such as choosing between oral antibiotics, a higher number of false positives may
be tolerable as long as treatment cost is not unbearably high. However, for highly
invasive and irreversibly treatments such as colectomy, false positives must be minimized.

bezlotoxumab shown to prevent systemic organ damage in a mouse study [@mileto_bezlotoxumab_2022].
-->
